Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-12-16 article.tex
Add ref to Strelkowa & Lassig

We focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one.
%If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fix.
The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments.
By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect a consistent tendency to rise in frequency (see below).

\subsection*{Predicting future frequencies}
Having observed the frequency trajectory $f(t)$ of a mutation until a given date $t_0$, how much can we say about the future values of $f$ after $t_0$?
We consider the idealized case sketched in panel \textbf{A} of figure~\ref{fig:trajectories}: given the trajectory of a \emph{new} mutation, \emph{i.e.} that started at a frequency of 0, and that we observe at frequency $f_0$ at time $t_0$, what is the probability $P_{\Delta t}(f)$ of observing it at a value $f$ at time $t_0 + \Delta t$?

>>>>

We focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one.
%If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fix.
The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments.
By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect a average \pierre{replaced ``consistent'' by ``average''} tendency to rise in frequency (see below). \pierre{Is this ``see below'' necessary? I think that's what the reader naturally does! }

\subsection*{Predicting future frequencies}
Having observed the frequency trajectory $f(t)$ of a mutation until a given date $t_0$, how much can we say about the future values of $f$ after $t_0$?
We consider the idealized case sketched in panel \textbf{A} of figure~\ref{fig:trajectories}: given the trajectory of a \emph{new} mutation, \emph{i.e.} that started at a frequency of 0, and that we observe at frequency $f_0$ at time $t_0$, what is the probability $P_{\Delta t}(f\vert f_0)$ of observing it at a value $f$ at time $t_0 + \Delta t$? 
This use of the frequency propagator to quantify the strength of selection was also proposed in \citep{strelkowa_clonal_2012}.
